---
title: "The 'memo' package"
author: "Peter Meilstrup"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The 'memo' package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
#memo
======
```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```
The `memo` package implements a simple in-memory cache for results of a function. If you have en expensive function that is being called repeatedly with the same inputs, `memo` can help.

## Fibonnacci example

```{r}
fib <- function(n) if (n <= 1) 1 else fib(n-1) + fib(n-2)
sapply(0:9, fib)
```

This recursive implementation corresponds closely to the way the sequence is defined in math texts, but it has a performance problem. The problem is that as you ask for values further down the sequence, the computation becomes inordinately slow due to all the recursion. To demonstrate the issue, we can try counting every time `fib` is
called:

```{r}
count <- 0
fib <- function(n) {
  count <<- count+1
  if (n <= 1) 1 else fib(n-1) + fib(n-2)
}

counted_fib <- function(n) {
  count <<- 0
  c(n=n, result=fib(n), calls=count)
}

t(sapply(0:16, counted_fib))
```

The number of calls increases unreasonably. This is because, for instance, `fib(6)` calls both `fib(5)` and `fib(4)`, but `fib(5)` also calls `fib(4)`. The second call to `fib(4)` is wasted work. This gets worse the higher `n` goes -- Every time you increment `n` by one, the number of calls roughly doubles.  (Clearly, there are more efficient algorithms for computing the Fibbonacci sequence, but this is a toy example, where `fib` stands in for some expensive function that is being called repeatedly.)

One way to cut down on wasted effort would be to check whether `fib(n)` has already been computed for a given `n`. If it has, `fib` can just return that value instead of starting over. This is called "memoizing." The `memo` package can [automatically]() create a memoized version of a given function, just by wrapping the function definition in `memo()`:

[automatically]:  https://en.wikipedia.org/wiki/Memoization#Automatic_memoization

```{r}
library(memo)

count <- 0
fib <- memo(function(n) {
  count <<- count+1
  if (n <= 1) 1 else fib(n-1) + fib(n-2)
})

counted_fib(16)
```
Now, computing `fib(16)` only takes 17 calls. And if we call again, it remembers the previous answer and doesn't make any new calls:
```{r}
counted_fib(16)
```
Each successive value then only takes two calls:
```{r}
t(sapply(17:30, counted_fib))
```

More practically, think of `memo` to make a cache for expensive database lookups, especially in multi-user Shiny apps where many users make identical queries. Memoization is also useful to optimize algorithms for path finding and recursive descent.
