---
title: "The \code{memo} package"
author: "Peter Meilstrup"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
#memo
======

The `memo` package implements a simple in-memory cache for the results
of repeated computations.

## Fibonnacci example

Consider this terrible way to compute the Fibonnacci sequence:

```{r}
fib <- function(n) if (n <= 1) 1 else fib(n-1) + fib(n-2)
```

Written this way, the function `fib` recursively invokes itself
many times, often to re-compute the same function many times over
again. We can see how many by counting calls in a global variable:

```{r}
count.calls <- function(f) {
  force(f)
  function(...) {
    count <<- count+1;
    f(...)
  }
}

with_count <- function(f) {
  force(f)
  function(...) {
    count <<- 0
    c(result=f(...), calls=count)
  }
}

fib <- count.calls(fib)

sapply(1:16, with_count(fib))
```

The number of calls increases unreasonably. This is because, say,
`fib(6)` needs to compute both `fib(5)` and `fib(4)`, but `fib(5)`
already computed `fib(4)`, so the second computation is wasted
work. We would be in better shape if later invocations of `fib` could
access data from earlier invocations.

By wrapping `fib` using `memo`:

```{r}
fib <- memo(fib)
sapply(1:16, with_count(fib))
```

Here the original function is only called to compute new values.
